How access to care shapes disease burden: the current impact of post-exposure prophylaxis and potential for expanded access to prevent human rabies deaths in Madagascar

### 2020-05-26 docs/manuscript.Rmd
cleaning up repo

---
title: "How access to care shapes disease burden: the current impact of post-exposure prophylaxis and potential for expanded access to prevent human rabies deaths in Madagascar"
author:
  - Malavika Rajeev:
      email: mrajeev@princeton.edu
      institute: [Princeton]
      correspondence: true
  - Helene Guis:
      institute: [IPMepi, CIRAD]
      correspondence: false
  - Glenn Edosoa:
      institute: [OMS]
      correspondence: false
  - Chantal Hanitriniaina:
      institute: [UoA]
      correspondence: false
  - Anjasoa Randrianarijaona:
      institute: [IPMepi]
      correspondence: false
  - Reziky Tiandraza Mangahasimbola:
      institute: [IPMepi]
      correspondence: false
  - Fleur Hierink:
      institute: [Geneva]
      correspondence: false
  - Ravo Ramiandrasoa:
      institute: [IPMvacc]
      correspondence: false
  - Jean-Michel Heraud:
      institute: [IPMviro]
      correspondence: false
  - Soa Fy Andriamandimby:
      institute: [IPMviro]
      correspondence: false
  - Laurence Baril:
      institute: [IPCambodia]
      correspondence: false
  - C.J.E. Metcalf:
      institute: [Princeton]
      correspondence: false
  - Katie Hampson:
      institute: [Glasgow]
      correspondence: false
institute:
  - Princeton: Department of Ecology and Evolutionary Biology, Princeton University, Princeton, United States
  - IPMviro: Virology Unit, Institut Pasteur de Madagascar, Antananarivo, Madagascar
  - IPMvacc: Vaccination Center, Institut Pasteur de Madagascar, Antananarivo, Madagascar
  - IPMepi: Epidemiology and Clinical Research Unit, Institut Pasteur de Madagascar, Antananarivo, Madagascar
  - IC: Institut Pasteur de Cambodge
  - CIRAD: CIRAD, UMR ASTRE, Antananarivo, Madagascar
  - OMS: OMS
  - UoA: Mention Zoologie et Biodiversité Animale, Faculté des Sciences, Université d’Antananarivo, Antananarivo, Madagascar
  - Glasgow: Boyd Orr Centre for Population and Ecosystem Health, Institute of Biodiversity, Animal Health and Comparative Medicine, University of Glasgow, Glasgow, UK
  - Geneva: University of Geneva
date: ""
link-citations: no
output:
 bookdown::word_document2:
    pandoc_args:
      - --filter=pandoc-crossref
      - --lua-filter=format/scholarly-metadata.lua
      - --lua-filter=format/author-info-blocks.lua
      - --lua-filter=format/pagebreak.lua
      - --lua-filter=format/multiple-bibliographies.lua
      - --reference-doc=format/Reference_Document.docx
 bookdown::html_document2:
    toc: true
    toc_float: true
    pandoc_args:
      - --csl=format/plos-neglected-tropical-diseases.csl
      - --filter=pandoc-crossref
      - --lua-filter=format/scholarly-metadata.lua
      - --lua-filter=format/author-info-blocks.lua
      - --lua-filter=format/pagebreak.lua
      - --lua-filter=format/multiple-bibliographies.lua
keywords: "canine rabies, post-exposure prophylaxis, Gavi, Zero by 30, access to care"
bibliography: ["bib/refs_all.bib", "bib/refs_pkgs.bib"]
# bibliography_main: ["bib/refs_all.bib"]
# bibliography_supp: ["bib/refs_pkgs.bib"]
---

##### TO DO: fix affiliations

```{r setupmain, include=FALSE}
# Libraries
library(knitr)
library(tidyverse)
library(data.table)
library(rgdal)
library(glue)
library(here)
library(magrittr)

opts_chunk$set(echo = FALSE, warning = FALSE, include = FALSE)

## data
ctar_metadata <- fread(here("data/raw/ctar_metadata.csv"))
district_bites <- fread(here("output/bites/district_bites.csv"))
mora_bites <- fread(here("output/bites/mora_bites.csv"))
mada_communes <- readOGR(here("data/processed/shapefiles/mada_communes.shp"))
mada_districts <- readOGR(here("data/processed/shapefiles/mada_districts.shp"))

@@

```
\newpage

# ABSTRACT
Access to care shapes burden of diseases globally. For dog-mediated rabies, post-exposure prophylaxis (PEP) is highly effective at preventing human deaths, however access to PEP is limited in rabies endemic countries. Recently, a target to eliminate deaths due to canine rabies has been set forth by global health partners. In 2018, Gavi added human rabies vaccine to its investment portfolio, a decision which has the potential to greatly reduce rabies deaths. Here, we estimate rabies burden in Madagascar incorporating estimates of geographic access to PEP.

Go in order.

We find that travel times to the closest Anti-Rabies Medical Center (ARMC) is a strong predictor of reported bite incidence across the country. We use resulting estimates of bite incidence in an adapted decision tree framework to generate estimates of burden and reporting across the country, and find that spatial access to PEP shapes local burden.

We evaluate the impact of expanding access to PEP, using a strategic framework which minimizes travel time for people at the greatest risk of death given a rabies exposure (persons > 3 hrs away from a clinic). Under these assumptions, we estimate that provisioning of PEP to an additional 200 clinics in Madagascar could reduce the burden of rabies deaths by 50%.

However our estimates are dependent on blah blah blah.

These results highlight the importance of accounting for access to care in predicting the impacts of interventions sub-nationally and the potential for expanding PEP access strategically to reduce rabies deaths in endemic settings.

# AUTHOR SUMMARY

@@

Dog-mediated rabies is estimated to cause approximately 60,000 human deaths annually [@whorabiesmodelingconsortiumPotentialImpactImproved2019]. Mass vaccination of domestic dogs has been demonstrated to be a highly effective way to control the disease in both animals and humans . While dog vaccination can interrput transmission in the reservoir, human deaths can also be prevented through prompt administration of post-exposure prophylactic vaccines (PEP) following a bite by a rabid dog [@RabiesVaccinesWHO2018]. However, access to human vaccines are limited in countries where canine rabies is endemic [@sreenivasanOverviewRabiesPostexposure2019; @tarantolaRabiesPostexposureProphylaxis2017; @hampsonEstimatingGlobalBurden2015], and within countries these deaths are often concentrated in rural, underserved communities [@wentworthSocialJusticePerspective2019].

In 2015, a global framework to eliminate deaths due to dog-mediated rabies by 2030 through a combination of PEP provisioning and dog vaccination was established by the World Health Organization (WHO) and partners [@world2018zero]. Furthermore, in 2018, Gavi, the Vaccine Alliance, added human rabies vaccines to their proposed investment portfolio [@GaviVIS]. From 2021, Gavi-eligible countries will be able to apply for support to expand access to these vaccines, with potential to greatly reduce deaths due to rabies. Yet it is unclear how these vaccines should be distributed to prevent deaths most effectively.

A primary challenge is the lack of data on rabies exposures in humans and incidence in animals in most rabies-endemic countries [@scottAddressingDisconnectEstimated2017]. The majority of rabies burden studies at the national level have integrated bite patient data on reported bites into a probability decision tree framework, with the key assumption that overall reported bite incidence (i.e. both bites due to non-rabid and rabid animals) are proportional to rabies incidence (i.e. the more bites reported in a location, the higher the incidence of rabies exposures), and that reporting is uniform across space [@hampsonEstimatingGlobalBurden2015; @knobelReevaluatingBurdenRabies2005; @rajeevHealthcareUtilizationProvisioning2018]. If applied subnationally, these assumptions would likely underestimate rabies deaths in places with poor access to PEP and may overestimate rabies deaths in places with better access to PEP.

@@

## Estimating travel time to closest ARMC {#methods-spatial}
To estimate travel times to the nearest clinic at the scale of administrative units, we used two raster inputs: 1) the friction surface from the Malaria Atlas Project [@weissGlobalMapTravel2018] at an ~1 km^2 scale ([Fig S1.1A](#figS1.1)) and 2) the population estimates which we resampled to the friction surface from the the 2015 UN adjusted population projections from World Pop (@linardPopulationDistributionSettlement2012a, originally at an ~100m^2 resolution, [Fig S1.1B](#figS1.1)).

@@

The data from the Moramanga district were completely reported across the `r max(mora_bites$nobs, na.rm = TRUE)` month period. However, for the national data, some clinics did not report any data, and for others there were substantial periods, months to even a whole year, when no patient forms were submitted. To correct for under-submission of forms, we attempted to identify what proportion of data were missing each year by excluding any periods of time for which there were 15 consecutive days where zero patient records were submitted ([Figure S2.1A](figS2.1). We then divided the total number of bites reported to each clinic from each clinic in a given year by the estimated proportion of forms which were not submitted (undersubmission) for each clinic. Due to the variation in submission of forms across time, we chose to take the average of the annual bite incidence estimates aggregated to the district level.

@@

```{r fig 1, fig.fullwidth=TRUE}
include_graphics(here("figs/main/M1_sankey.jpeg"))
```
*__Figure 2.__Decision tree for burden estimation. For a given administrative unit $i$, human deaths due to rabies ($D_i$) are calculated from model predicted bites ($B_i$). To get $R_i$, the number of reported bites that were rabies exposures, we multiply $B_i$ by $p_{rabid}$, the proportion of reported bites that are rabies exposures. $p_{rabid}$ is constrained such that reported rabies exposures cannot exceed total rabies exposures ($E_i$) and reporting cannot exceed a maximum ($p_{max}$). $R_i$ is subtracted from $E_i$ to get the number of unreported bites ($U_i$) and then multiplied by the probability of death given a rabies exposure ($p_{infect}$) to get the number of deaths due to rabies ($D_i$). Similarly, deaths averted by PEP are estimated by multiplying $R_i$ by $p_{infect}$, i.e. those who would have died given exposure, but instead received PEP. Both $E_i$ and $p_{rabid}$ are drawn from a triangular distribution; all parameter values and sources are listed in Table 1.*

@@

```{r ncsbs}
## candidate points
csbs <- fread(here("data/processed/clinics/csb2.csv"))
clin_per_comm <- fread(here("data/processed/clinics/clinic_per_comm.csv"))
clin_per_dist <- fread(here("data/processed/clinics/clinic_per_dist.csv"))
missing_comms <- fread(here("output/stats/missing_comms.csv"))
comms_no_clins <- length(mada_communes) - (nrow(clin_per_comm) + nrow(ctar_metadata) + nrow(clin_per_dist))
csbs_max <- rbind(csbs, clin_per_comm[!(clinic_id %in% csbs$clinic_id)], fill = TRUE)

@@

## Data and code 
All analyses were done in `r sessionInfo()$R.version$version.string` [@RLang2019] and using the packages listed in the [supplementary references]{#srefs}. All processed data, code, and outputs are archived at this zenodo repository (to be archived upon submission) and maintained at [this github repository](https://github.com/mrajeev08/MadaAccess). The raw bite patient data at the national level are maintained in a database by IPM. This database was last queried on September 19, 2019 for these analyses. Anonymized raw data from this database are available from IPM following insitutional data transfer protocols. Anonymized raw data collected from the Moramanga District were retrieved from the Wise Monkey Portal (http://www.wisemonkeyfoundation.org/) on the same date and are shared in the archived repository.

@@

Estimates of travel times to the closest ARMC vary greatly across the country (Fig 3A), with approximately `r round(prop_over3_base*100, 0)`% of the population living greater than 3 hrs away from a clinic (Fig 3B). We found that travel times estimated from the friction surface generally underestimate true driving times and patient reported travel times to the Moramanga ARMC (Fig 3C), and that patient reported travel times were significantly more variable than estimates from the friction surface (Fig S1.2). The friction surface assumes that the fastest available mode of transport is used to travel each grid cell (i.e. the presence of a road in a given grid cell indicates that all travel through that grid cell will be at vehicle rather than walking speeds). However, patients reporting to the Moramanga ARMC report using mutliple modes of transport, with some cases of individuals walking multiple days to reach a clinic (Fig S1.3). Despite these discrepancies, travel times correlate with the reported and groundtruthed driving times (Fig 3C, Fig S1.4). Travel times weighted by population at the grid cell level were a better predicter than unweighted travel times or distance (Table S1.1). Therefore we use this metric as a proxy for access at the commune/district level in subsequent analyses.

```{r fig M2, fig.fullwidth=TRUE}
include_graphics(here("figs/main/M2_ttimes.jpeg"))
```
*__Figure 3.A)__ Travel times to the closest ARMC estimated at an ~ 1km^2 scale for Madagascar. __B)__ Distribution of the population across travel times. __C)__Correlation between patient reported travel times to the Moramanga ARMC and travel times weighted by population (mean at the commune level) and between reported driving times between GPS points and travel time estimates at the grid cell level. For the commune means, the vertical point lineranges show the 95% quantiles for reported travel times, and the point size shows the number of observations for each commune. The best fit lines (pink and purple lines) and 95% confidence interval (shading) from a linear model where observed travel times ~ estimated travel times for each data source are also shown. The dashed black line is the 1:1 line.*

@@

```{r fig M3 catchmaps}
include_graphics(here("figs/main/M3_bitemaps.jpeg"))
```
*__Figure 3.A)__ The network of patient presentations at the district level for the __A)__ national data and __B)__ commune level for the Moramanga data: circles with a black outline represent the total number of patients reporting to each ARMC for which we have data. Color corresponds to the clinic catchment. Circles with a white outline are the total number of bites reported for that administrative unit (plotted as the centroid). Lines show which ARMC those patients reported to, with the line width proportional to number of patients from that district reporting to the ARMC; flows of less than 5 patients were excluded. Out-of-catchment reporting is indicated where points and line colors are mismatched. For panel __A)__ districts in catchments excluded due to lack of forms submitted by the clinic are colored in grey. For __B)__ the inset of Madagascar shows the location of the enlarged area plotted, which shows the district of Moramanga (outlined in black), and all communes included in it's catchment (red polygons) and other communes where bites were reported to in grey.*

Clinic submission of forms was highly variable over the period covered by the dataset (Figure S2.1). We found that estimates of vial demand based on uncorrected bite patient numbers are generally lower than the true number of vials provisioned for those clinics with substantial undersubmission of forms ([Figure S2.1B](#figS2.1)), and correcting patient numbers by under-submission of forms based on the 15 day consecutive day cut-off results in estimates of vial demand closer to the observed vials provisioned ([Figure S2.2A and B](#figS2.2)). We excluded any clinics which submitted less than 10 forms (excluded `r nrow(ctar_metadata) - max(district_bites$catch)` catchments, in grey in Fig 3A). After additionally excluding any year with less than 25% of forms submitted, our final dataset consisted of estimates of average bite incidence for `r nrow(district_bites)` out for `r length(mada_districts)` districts total (Fig 4A), and `r nrow(mora_bites)` communes within the catchment of the Moramanga District (Fig 4B).

@@

Bite incidence estimates generally decreased with weighted travel times at both scales, although there was considerable variation between catchments for the magnitude of this relationship. For the national data, there were also two outliers, Toamasina II (the sub-urban commune surrounding the city of Toamasina) and Soanierana Ivongo, which both had higher bite incidence when compared to other districts with similar travel times. While the estimates from the Moramanga data showed higher reported incidence at low travel times at the commune level compared to the district estimates, when aggregated to the district fell within the ranges observed from the national dataset.

@@

For the national data, including a random catchment effect improved predictions of the observed data (Fig S3.2 & Fig S3.3), and catchment level effects were strongly identified in the posterior estimates. However, after accounting for overdispersion of the data, these catchment effects were not clearly identifiable and the posterior estimates of the catchment intercepts were broadly similar (Fig S3.6), and the models resulted in similar predictions to the data (Fig S3.7), indicating that the catchment effect could not be differentiated from random variation in the data. Similarly, while the commune model fit to the Moramanga data generated stronger estimates of the travel time effect (Fig 5B), after accounting for overdispersion of the data, the posterior estimates of the parameters overlapped for the commune and district models fit to the national data (Fig S3.4), and the model estimates were in general less robust to overdispersion than for the national data, particularly at low travel times (Fig S3.5). Finally, for the national data, we also looked at the sensitivity of these models to how we corrected our data for underreporting, and we found that parameter estimates of travel time impacts were similar across models (Fig S3.8) and generated similar predictions of the range of bite incidence across travel times (Fig S3.9).

As the prediction from the model fit to the Moramanga data without accounting for overdispersion falls within the prediction interval for the models fit to the national data (Fig 5A), for subsequent predictions of burden, we use the parameter estimates from the commune and district models fit to the national data, as they encompass the range of travel time effects observed in our datasets.

>>>>

---
title: "How access to care shapes disease burden: the current impact of post-exposure prophylaxis and potential for expanded access to prevent human rabies deaths in Madagascar"
author:
  - Malavika Rajeev:
      email: mrajeev@princeton.edu
      institute: [Princeton]
      correspondence: true
  - Helene Guis:
      institute: [IPMepi, CIRAD]
      correspondence: false
  - Glenn Edosoa:
      institute: [OMS]
      correspondence: false
  - Chantal Hanitriniaina:
      institute: [UoA]
      correspondence: false
  - Anjasoa Randrianarijaona:
      institute: [IPMepi]
      correspondence: false
  - Reziky Tiandraza Mangahasimbola:
      institute: [IPMepi]
      correspondence: false
  - Fleur Hierink:
      institute: [Geneva]
      correspondence: false
  - Ravo Ramiandrasoa:
      institute: [IPMvacc]
      correspondence: false
  - Jean-Michel Heraud:
      institute: [IPMviro]
      correspondence: false
  - Soa Fy Andriamandimby:
      institute: [IPMviro]
      correspondence: false
  - Laurence Baril:
      institute: [IC]
      correspondence: false
  - C.J.E. Metcalf:
      institute: [Princeton]
      correspondence: false
  - Katie Hampson:
      institute: [Glasgow]
      correspondence: false
institute:
  - Princeton: Department of Ecology and Evolutionary Biology, Princeton University, Princeton, United States
  - IPMepi: Epidemiology and Clinical Research Unit, Institut Pasteur de Madagascar, Antananarivo, Madagascar
  - CIRAD: CIRAD, UMR ASTRE, Antananarivo, Madagascar
  - OMS: World Health Organization, Madagascar
  - UoA: Mention Zoologie et Biodiversité Animale, Faculté des Sciences, Université d’Antananarivo, Antananarivo, Madagascar
  - Geneva: University of Geneva
  - IPMvacc: Vaccination Center, Institut Pasteur de Madagascar, Antananarivo, Madagascar
  - IPMviro: Virology Unit, Institut Pasteur de Madagascar, Antananarivo, Madagascar
  - IC: Institut Pasteur de Cambodge
  - Glasgow: Boyd Orr Centre for Population and Ecosystem Health, Institute of Biodiversity, Animal Health and Comparative Medicine, University of Glasgow, Glasgow, UK
date: ""
link-citations: no
nocite: '@*'
output:
 bookdown::word_document2:
    pandoc_args:
      - --csl=format/plos-neglected-tropical-diseases.csl
      - --filter=pandoc-crossref
      - --lua-filter=format/scholarly-metadata.lua
      - --lua-filter=format/author-info-blocks.lua
      - --lua-filter=format/pagebreak.lua
      - --lua-filter=format/multiple-bibliographies.lua
      - --reference-doc=format/Reference_Document.docx
 bookdown::html_document2:
    toc: true
    toc_float: true
    number_sections: no
    pandoc_args:
      - --csl=format/plos-neglected-tropical-diseases.csl
      - --filter=pandoc-crossref
      - --lua-filter=format/scholarly-metadata.lua
      - --lua-filter=format/author-info-blocks.lua
      - --lua-filter=format/pagebreak.lua
      - --lua-filter=format/multiple-bibliographies.lua
keywords: "canine rabies, post-exposure prophylaxis, Gavi, Zero by 30, access to care"
bibliography_main: ["bib/refs_all.bib"]
bibliography_supp: ["bib/refs_pkgs.bib"]
---

```{r setupmain, include=FALSE}
# Libraries
library(knitr)
library(tidyverse)
library(data.table)
library(rgdal)
library(glue)
library(here)
library(magrittr)
opts_chunk$set(echo = FALSE, warning = FALSE, out.width="80%")

@@

## data
ctar_metadata <- fread(here("data/processed/clinics/ctar_metadata.csv"))
district_bites <- fread(here("output/bites/district_bites.csv"))
mora_bites <- fread(here("output/bites/mora_bites.csv"))
mada_communes <- readOGR(here("data/processed/shapefiles/mada_communes.shp"))
mada_districts <- readOGR(here("data/processed/shapefiles/mada_districts.shp"))

## All preds (for abstract stats etc)
natl_preds <- fread(here("output/preds/natl_preds.gz"))
natl_base <- natl_preds[scenario == 0]

natl_preds %>%
  filter(scenario == 0) %>%
  mutate(prop_averted = averted_mean/(deaths_mean + averted_mean)) %>%
  select(prop_averted) -> prop_averted

admin_base <- fread(here("output/preds/admin_preds_base.gz"))

## candidate points
csbs <- fread(here("data/processed/clinics/csb2.csv"))
clin_per_comm <- fread(here("data/processed/clinics/clinic_per_comm.csv"))
clin_per_dist <- fread(here("data/processed/clinics/clinic_per_dist.csv"))
missing_comms <- fread(here("output/stats/missing_comms.csv"))
comms_no_clins <- length(mada_communes) - (nrow(clin_per_comm) + nrow(ctar_metadata) + nrow(clin_per_dist))
csbs_max <- rbind(csbs, clin_per_comm[!(clinic_id %in% csbs$clinic_id)], fill = TRUE)

max_added <- max(natl_preds$scenario)
natl_preds %>%
  group_by(scale, data_source) %>%
  arrange(scenario) %>%
  mutate_at(vars(-c(group_cols(), scenario)), function(x) x/x[1]) -> props
trends <- read.csv(here("output/stats/trends_adminpreds.csv"))

one_per_dist <- nrow(clin_per_dist)
one_per_comm <- sort(natl_preds$scenario, decreasing = TRUE)[2]

no_figs <- FALSE
if(no_figs) include_graphics <- function(x, ...) { return(invisible(NULL)) }

# ABSTRACT

## Background
For dog-mediated rabies, post-exposure prophylaxis (PEP) is highly effective at preventing human deaths, however access to PEP is limited in rabies endemic countries. A target to eliminate deaths due to canine rabies by the year 2030 has been set forth by global health partners, and in 2018, Gavi added human rabies vaccine to its investment portfolio, a decision which has the potential to greatly expand PEP availability and reduce rabies burden. Given the potential for countries to provision PEP more widely, we estimate the impact of access to PEP on burden and explore how expanding access can reduce burden in Madagascar, where current access to PEP is highly limited spatially (N = 31 Anti-Rabies Medical Center or ARMC across the country).

## Methodology & Principal Findings
We use travel times to the closest ARMC as a proxy for access to PEP. We find that travel times are a strong predictor of reported bite incidence across the country. We use resulting estimates of bite incidence in an adapted decision tree framework to generate estimates of burden and reporting across the country, and find that spatial access to PEP shapes burden sub-nationally. We estimate an annual burden of `r round(sort(natl_base$deaths_mean)[1])` (95% CI: `r round(sort(natl_base$deaths_lower)[nrow(natl_base)])` - `r round(sort(natl_base$deaths_upper)[nrow(natl_base)])`), with current PEP provisioning averting ~ `r glue("{trans_perc(min(prop_averted))}%")` percent of deaths. Under these assumptions, we estimate that expanding PEP provisioning to one clinic per district could reduce the burden of rabies deaths by `r glue("{trans_perc(min(1 - props$deaths_mean[props$scenario == 83]))} - {trans_perc(max(1 - props$deaths_mean[props$scenario == 83]))}%")`. However, our results suggest that even with all major health centers in the country provisioning PEP (N = `r max(natl_preds$scenario)` additional clinics), we could still see substantial mortality due to rabies. Our estimates are most sensitive to assumptions of underlying rabies exposure incidence, but the qualitative patterns of the impacts of travel times and expanded access on burden are robust to these assumptions.

## Conclusions & Significance
PEP is effective at preventing deaths and in the absence of rabies data, targeting expansion to populations currently underserved may be the most equitable way to provision PEP. Data on rabies incidence will be neccessary to refine these estimates and to better target interventions. While expanding access could save many more lives, our results suggest that improved outreach and awareness is also needed & improved clinic functioning & surveillance systems. A GAVI investment can propel countries forward on this path combining these life saving interventions.

# AUTHOR SUMMARY 
Canine rabies causes an estimated 60,000 deaths each year across the world, primarily in low and middle income countries where people have limited access to both human (post-exposure prophylaxis or PEP) and dog rabies vaccine. Given that we have the tools to prevent rabies deaths, a global target has been set to eliminate deaths due to canine rabies by the year 2030, and recently, GAVI, a organization that provides funding for immunization programs to eligible countries, added human rabies vaccine to it's funding agenda. In this study, we model estimates of reported bite incidence as a function of travel times and estimates of rabies exposure incidence to extrapolate the burden of human rabies given the current availability of PEP across the country. We find that PEP currently averts a signficant number of deaths, but that burden remains high, particularly in areas with high travel times to a clinic. Expanding PEP availability to more clinics could significantly reduce the current burden of rabies deaths, but our results suggest that expansion alone will not eliminate deaths due to rabies under current levels of reporting. Combining PEP expansion with outreach, surveillance, and mass dog vaccination programs will be necessary to move Madagascar, and other countries like it, forwards on the path to elimination.

@@

Dog-mediated rabies is estimated to cause approximately 60,000 human deaths annually [@whorabiesmodelingconsortiumPotentialImpactImproved2019]. Mass vaccination of domestic dogs has been demonstrated to be a highly effective way to control the disease in both animals and humans. While dog vaccination can interrput transmission in the reservoir, human deaths can also be prevented through prompt administration of post-exposure prophylactic vaccines (PEP) following a bite by a rabid dog [@RabiesVaccinesWHO2018]. However, access to human vaccines are limited in countries where canine rabies is endemic [@sreenivasanOverviewRabiesPostexposure2019; @tarantolaRabiesPostexposureProphylaxis2017; @hampsonEstimatingGlobalBurden2015], and within countries these deaths are often concentrated in rural, underserved communities [@wentworthSocialJusticePerspective2019].

In 2015, a global framework to eliminate deaths due to dog-mediated rabies by 2030 through a combination of PEP provisioning and dog vaccination was established by the World Health Organization (WHO) and partners [@world2018zero]. Furthermore, in 2018, Gavi, the Vaccine Alliance, added human rabies vaccines to their proposed investment portfolio [@GaviVIS]. From 2021, Gavi-eligible countries will be able to apply for support to expand access to these vaccines, with potential to greatly reduce deaths due to rabies.

A primary challenge in expanding access effectively is the lack of data on rabies exposures and deathss in humans and incidence in animals in most rabies-endemic countries [@scottAddressingDisconnectEstimated2017]. Deaths due to rabies are often severely underreported, with many people dying outside of the health system, often in the most vulnerable and remote communities [@taylor2017difficulties]. Instead of directly measuring rabies deaths, the majority of rabies burden studies use bite patient data on reported bites and a decision tree framework to extrapolate deaths, with the key assumption that overall reported bite incidence (i.e. both bites due to non-rabid and rabid animals) are proportional to rabies incidence (i.e. the more bites reported in a location, the higher the incidence of rabies exposures), and that reporting is uniform across space [@hampsonEstimatingGlobalBurden2015; @knobelReevaluatingBurdenRabies2005; @rajeevHealthcareUtilizationProvisioning2018]. If applied subnationally, these assumptions would likely underestimate rabies deaths in places with poor access to PEP and may overestimate rabies deaths in places with better access to PEP.

@@

## Estimating travel time to closest ARMC {#methods-ttimes}
To estimate travel times to the nearest clinic at the scale of administrative units, we used two raster inputs: 1) the friction surface from the Malaria Atlas Project [@weissGlobalMapTravel2018] at an ~1 km^2 scale (Fig S1.1A) and 2) the population estimates which we resampled to the friction surface from the the 2015 UN adjusted population projections from World Pop (@linardPopulationDistributionSettlement2012a, originally at an ~100m^2 resolution, Fig S1.1B.

@@

The data from the Moramanga district were completely reported across the `r max(mora_bites$nobs, na.rm = TRUE)` month period. However, for the national data, some clinics did not report any data, and for others there were substantial periods, months to even a whole year, when no patient forms were submitted. To correct for under-submission of forms, we attempted to identify what proportion of data were missing each year by excluding any periods of time for which there were 15 consecutive days where zero patient records were submitted ([Figure S2.1A](figS2.1)). We then divided the total number of bites reported to each clinic from each clinic in a given year by the estimated proportion of forms which were not submitted (undersubmission) for each clinic. Due to the variation in submission of forms across time, we chose to take the average of the annual bite incidence estimates aggregated to the district level.

@@

```{r fig 1}
include_graphics(here("figs/main/M1_sankey.jpeg"))
```

*__Figure 1__. Decision tree for burden estimation. For a given administrative unit $i$, human deaths due to rabies ($D_i$) are calculated from model predicted bites ($B_i$). To get $R_i$, the number of reported bites that were rabies exposures, we multiply $B_i$ by $p_{rabid}$, the proportion of reported bites that are rabies exposures. $p_{rabid}$ is constrained such that reported rabies exposures cannot exceed total rabies exposures ($E_i$) and reporting cannot exceed a maximum ($p_{max}$). $R_i$ is subtracted from $E_i$ to get the number of unreported bites ($U_i$) and then multiplied by the probability of death given a rabies exposure ($p_{infect}$) to get the number of deaths due to rabies ($D_i$). Similarly, deaths averted by PEP are estimated by multiplying $R_i$ by $p_{infect}$, i.e. those who would have died given exposure, but instead received PEP. Both $E_i$ and $p_{rabid}$ are drawn from a triangular distribution; all parameter values and sources are listed in Table 1.*

@@

@@

## Data and analyses 
All analyses were done in `r sessionInfo()$R.version$version.string` [@RLang2019] and using the packages listed in the supplementary references (section S7). All processed data, code, and outputs are archived at this zenodo repository (to be archived upon submission) and maintained at github.com/mrajeev08/MadaAccess. The raw bite patient data at the national level are maintained in two secure REDCap (project-redcap.org) databases, one for IPM and another for all peripheral ARMC. These databases were last queried on September 19, 2019 for these analyses. Anonymized raw bite patient data are available from IPM following insitutional data transfer protocols. Anonymized raw data collected from the Moramanga District were retrieved from the Wise Monkey Portal (wisemonkeyfoundation.org) on the same date and are shared in the archived repository.

@@

Estimates of travel times to the closest ARMC vary greatly across the country (Fig 3A), with approximately `r round(prop_over3_base*100, 0)`% of the population estimated to live greater than 3 hrs away from a clinic (Fig 3B). However, we found that these estimates from the friction surface generally underestimate true driving times across the country and patient reported travel times to the Moramanga ARMC (Fig 3C), and that patient reported travel times were significantly more variable than estimates from the friction surface (Fig S1.2). The friction surface assumes that the fastest available mode of transport is used to travel each grid cell (i.e. the presence of a road in a given grid cell indicates that all travel through that grid cell will be at vehicle rather than walking speeds). However, patients reporting to the Moramanga ARMC report using mutliple modes of transport, with some cases of individuals walking multiple days to reach a clinic (Fig S1.3). Despite these discrepancies, travel time estimates from the friction surface correlate with the reported and groundtruthed driving times (Fig 3C, Fig S1.4). Travel times weighted by population at the grid cell level were a better predicter than unweighted travel times or distance (Table S1.1). Therefore we use this metric as a proxy for access at the commune/district level in subsequent analyses.

```{r fig M2}
include_graphics(here("figs/main/M2_ttimes.jpeg"))
```

*__Figure 2A)__ Travel times to the closest ARMC estimated at an ~ 1 km^2^ scale for Madagascar. __B)__ Distribution of the population across travel times. __C)__ Correlation between patient reported travel times to the Moramanga ARMC and travel times weighted by population (mean at the commune level) and between reported driving times between GPS points and travel time estimates at the grid cell level. For the commune means, the vertical point lineranges show the 95% quantiles for reported travel times, and the point size shows the number of observations for each commune. The best fit lines (pink and purple lines) and 95% confidence interval (shading) from a linear model where observed travel times are predicted by estimated travel times for each data source are also shown. The dashed black line is the 1:1 line.*

@@

```{r fig M3 catchmaps}
include_graphics(here("figs/main/M3_bitemaps.jpeg"))
```

*__Figure 3A)__ The network of patient presentations at the district level for the __A)__ national data and __B)__ commune level for the Moramanga data: circles with a black outline represent the total number of patients reporting to each ARMC for which we have data. Color corresponds to the clinic catchment. Circles with a white outline are the total number of bites reported for that administrative unit (plotted as the centroid). Lines show which ARMC those patients reported to, with the line width proportional to number of patients from that district reporting to the ARMC; flows of less than 5 patients were excluded. Out-of-catchment reporting is indicated where points and line colors are mismatched. For panel __A)__ districts in catchments excluded due to lack of forms submitted by the clinic are colored in grey. For __B)__ the inset of Madagascar shows the location of the enlarged area plotted, which shows the district of Moramanga (outlined in black), all communes included in it's catchment (red polygons), and other communes where bites were reported to in grey.*

Clinic submission of forms was highly variable over the period covered by the dataset (Figure S2.1). We found that estimates of vial demand based on uncorrected bite patient numbers are generally lower than the true number of vials provisioned for those clinics with substantial undersubmission of forms (Figure S2.1B). Correcting patient numbers by under-submission of forms based on the 15 day consecutive day cut-off results in estimates of vial demand closer to the observed vials provisioned for the majority of clinics (Figure S2.2A and B). We excluded any clinics which submitted less than 10 forms (excluded `r nrow(ctar_metadata) - max(district_bites$catch)` catchments, in grey in Fig 3A). After additionally excluding any year with less than 25% of forms submitted, our final dataset consisted of estimates of average bite incidence for `r nrow(district_bites)` out for `r length(mada_districts)` districts total (Fig 4A), and `r nrow(mora_bites)` communes within the catchment of the Moramanga District (Fig 4B).

@@

Bite incidence estimates generally decreased with weighted travel times at both scales, although there was considerable variation between catchments for the magnitude of this relationship. For the national data, there were also two outliers, Toamasina II (the sub-urban commune surrounding the city of Toamasina) and Soanierana Ivongo, which both had higher bite incidence when compared to other districts with similar travel times. While the estimates from the Moramanga data showed higher reported incidence at low travel times at the commune level compared to the district estimates, when aggregated to the district, bite incidence estimates fell within the ranges observed from the national dataset.

@@

For the national data, including a random catchment effect improved predictions of the observed data (Fig S3.2 & Fig S3.3), and catchment level effects were strongly identified in the posterior estimates. However, after accounting for overdispersion of the data, these catchment effects were not clearly identifiable and the posterior estimates of the catchment intercepts were broadly similar (Fig S3.6), and the models resulted in similar predictions to the data (Fig S3.7), indicating that the catchment effect could not be differentiated from random variation in the data. Similarly, while the commune model fit to the Moramanga data generated stronger estimates of the travel time effect (Fig 5B), after accounting for overdispersion of the data, the posterior estimates of the parameters overlapped for the commune and district models fit to the national data (Fig S3.4), and the model estimates were in general less robust to overdispersion than for the national data, particularly at low travel times (Fig S3.5). Finally, for the national data, we also looked at the sensitivity of these models to how we corrected our data for underreporting, and we found that parameter estimates of travel time impacts were similar across models and performed similarly in predicting the data, and instead the correction applied impacted the magnitude of the bite incidence estimates (Fig S3.8 and Fig S3.9).

As the prediction from the model fit to the Moramanga data without accounting for overdispersion falls within the prediction interval for the models fit to the national data (Fig 5A), for subsequent predictions of burden, we use the parameter estimates models fit to the national data, as they encompass the range of travel time effects observed in our datasets. In addition, our out of fit predictions to the data across scales suggest that the commune model is able to capture travel time impacts at this scale, therefore we use both the district and commune model to disaggregate burden to the finest scale possible given our modeling framework.

```{r fig 5}
include_graphics(here("figs/main/M5_mods.jpeg"))
```
